Mathematical Aspects of the Capital Market

### 2017-07-21 phd/Chapterextremes.tex
Added my own thesis.

\begin{abstract}{We  provide some  asymptotic theory for the largest eigenvalues of a sample covariance matrix
of a $p$-dimensional time series where the dimension $p=p_n$ converges to infinity when the sample size $n$ increases.
We give a short overview of the literature on the topic both in the light- and heavy-tailed cases when the data have
finite (infinite) fourth moment, respectively.
Our main focus is on the heavy-tailed case. In this case, one has a theory for the point process of the normalized eigenvalues
of the sample covariance matrix in the iid case but also when rows and columns of the data are linearly dependent.
We provide limit results for the weak convergence of these point processes to Poisson or cluster Poisson processes. Based o
this convergence we can also derive the limit laws of various functionals of the ordered eigenvalues such as the
joint convergence of a finite number of the largest order statistics, the joint limit law of the largest eigenvalue and the trace,
limit laws for successive ratios of ordered eigenvalues,
etc. We also develop some
limit theory for the singular values of the sample autocovariance matrices and their sums of squares. The theory is illustrated
for simulated data and for the components of the S\&P 500 stock index.
\medskip

\noindent {\bf Keywords:} Regular variation, sample covariance matrix, dependent entries,
largest  eigenvalues, trace, point process
  convergence, cluster Poisson limit,
infinite variance stable limit, Fr\'echet distribution.}
\end{abstract}

\newpage
\section[Estimation of the largest eigenvalues]{Estimation of the largest eigenvalues: an overview in the iid case}\label{sec:motivationch3}
\subsection{The light-tailed case}
%In recent years, \asy\ theory for large random matrices has a
%attracted a
%great deal of attention; see for example the monographs
%Bai and Silverstein \cite{bai:silverstein:2010} and Anderson et al.
%\cite{anderson:guionnet:zeitouni:2008}. The literature about {\em
%  heavy-tailed random matrices} is rather sparse. Here and in what
%follows, we call a random matrix {\em heavy-tailed} if its entries have
%\ds s with \regvary\ tails, typically with tail index below 4.
%The eigenvalues of heavy-tailed matrices
%with iid entries and dimension $p\times n$, where $p=p_n\to\infty$ and $p/n\to
%\gamma\in (0,\infty)$, were studied by Soshnikov \cite{soshnikov:2004,
%  soshnikov:2006} and  Auffinger et
%al. \cite{auffinger:arous:peche:2009}. In the latter reference and in
%Belinschi et al. \cite{belinschi:dembo:guionnet:2009},
%the sample covariance matrices of iid heavy-tailed \seq s were also studied.
%Bose et al. \cite{bose:hazra:saha:2009,bose:hazra:saha:2010}
%investigated the spectral norm of circulant-type heavy-tailed matrices
%and Ben Arous and Guionnet \cite{arous:guionnet:2008}
%studied the limits of heavy-tailed random Wigner matrices with
%infinite variance.
One of the exciting new areas of statistics is concerned with analyses of large data sets.
For such data one often studies the dependence structure via covariances and correlations.
In this paper we focus on one aspect: the estimation of the eigenvalues of the covariance matrix of a multivariate  \ts\
when the dimension $p$ of the series increases with the sample size $n$. In particular, we are interested in limit theory for the largest eigenvalues of the sample covariance matrix. This theory is closely related to topics from classical \evt\ such as
maximum domains of attraction with the corresponding normalizing and centering constants for maxima; cf. Embrechts et al.~\cite{embrechts:kluppelberg:mikosch:1997}, Resnick~\cite{resnick:2007,resnick:1987}.
Moreover, \pp\ \con\ with limiting Poisson and cluster Poisson processes enters in a natural way when one describes
the joint \con\ of the largest eigenvalues of the sample covariance matrix. Large deviation techniques find applications,
linking \evt\ with random walk theory and \pp\ \con . The objective of this paper is to illustrate some of the main
developments in random matrix theory for the particular case of the sample covariance matrix of multivariate \ts\ with
independent or dependent entries. We give special emphasis to the heavy-tailed case when \evt\ enters in a rather
straightforward way.
%We do not cover other parts of random matrix theory such as limit theory for the eigen-spectrum of random matrices with
%iid entries or other special types of matrices. Also in this case one can find many parallels with \evt.
\par
Classical multivariate \tsa\ deals with
observations which assume values in a $p$-dimensional space
where $p$ is ``relatively small'' compared to the sample size $n$.  With
the availability of large data sets $p$ can be ``large''
relative to $n$. One of the possible con\seq s is that standard asymptotics (such as the \clt )
break down and may even cause misleading results.
\par
The dependence
structure in multivariate data is often summarized by the covariance matrix which is typically
estimated by its sample analog.  For example, principal component analysis (PCA)
extracts principal component vectors corresponding to the largest
eigenvalues of the sample covariance matrix. The magnitudes of these eigenvalues provide an empirical \ms\ of the importance
of these components.
\par
If $p,n$ are fixed, a column of the $p\times n$ data matrix
\beao
\X =\X_n = \big(X_{it}\big)_{i=1,\ldots,p;t=1,\ldots,n}\,
\eeao
represents an observation of a $p$-dimensional \ts\ model with unknown parameters.
In this section we assume that the real-valued entries $X_{it}$ are iid, unless mentioned otherwise, and we write $X$ for a generic element.
One challenge is to infer information about the parameters
from the eigenvalues $\lambda_1,\ldots,\lambda_p$
of the {\em sample covariance matrix} $\X\X'$. In the notation we suppress the dependence of $(\la_i)$ on $n$ and $p$.
If $p$ and $n$ are finite and the columns of $\X$ are iid and multivariate normal,
Muirhead \cite{muirhead} derived a (rather complicated) formula for the joint distribution of the eigenvalues $(\lambda_i)$.
%Therefore a theoretical solution is already available in the finite $n$ case.
%The expression, however, contains an integral over the orthogonal group $\mathbb{O}(p)$.
%\begin{figure}[htb!]
%  \centering
 % \includegraphics[scale=0.5]{eigen_normal_qqplot1.pdf}
 % \caption{Largest eigenvalues of $500$ simulated sample covariance matrices against standard normal quantiles.}
 % \label{fig:QQ}
%\end{figure}

For $p$ fixed and $\nto$, assuming
$\X$ has centered normal entries and a diagonal covariance matrix $\Sigma$,
Anderson~\cite{anderson:1963}  derived the joint \asy\ density of $(\lambda_1, \ldots, \lambda_p)$.
%If the largest eigenvalue of the underlying covariance matrix appears with multiplicity one, he
%showed that the largest eigenvalue of the sample covariave matrix
%is asymptotically normal.
We quote from Johnstone~\cite{johnstone:2001}:
``The classic paper by Anderson~\cite{anderson:1963} gives the limiting joint distribution of the roots, but the
marginal distribution of the largest eigenvalue is hard to extract even in the null case'' (i.e., when the covariance matrix $\Sigma$ is
proportional to the identity matrix).
%If the largest entry of $\Sigma$ exceeds its second largest, one obtains a central limit type theorem for the largest eigenvalue of the sample covariance matrix.
%In Figure~\ref{fig:QQ} we chose $n=10000, p=50$ and $\Sigma =\diag(1024,1,\ldots,1)$, and plot the properly normalized and
%standardized largest eigenvalue against standard normal quantiles.

It turns out that limit theory for the largest eigenvalues becomes ``easier''
when the dimension $p$ increases with $n$.
Over the last 15 years there has been increasing interest in the case when $p=p_n\to\infty$ as $\nto$. In most of
the literature (exceptions are El Karoui \cite{elkaroui:2003}, Davis et al. \cite{davis:mikosch:pfaffel:2015,davis:pfaffel:stelzer:2014} and Heiny and Mikosch
\cite{heiny:mikosch:2015:iid})
one assumes that $p$ and $n$ grow at the same rate:
\beam\label{eq:gammach3}
\dfrac p n\to \gamma\qquad \mbox{for some $\gamma \in (0,\infty)$.}
\eeam
%One explanation for the commonly used condition \eqref{eq:gamma} is that much of the theory was developed for $n \times n$ {\em Wigner matrices}, which is a Hermitean matrix $H$ such that $\{ H_{ij}:i<j \}$ and $\{ H_{ii}:0\le i\le n \}$ are two sets of iid random variables. If $p =\gamma n$, it turns out that the limit theory for the eigenvalues of Wigner matrices and sample covariance matrices is closely connected by the convergence of the empirical spectral distributions to the semicircle and the Mar\v cenko--Pastur law, respectively; see Bai and Silverstein \cite{bai:silverstein:2010}, Chapters 1-3.

In random matrix theory,  the convergence of the {\em empirical spectral distributions} $(F_{n^{-1}\X \X'})$ of a sequence $(n^{-1}\X \X')$ of non-negative definite matrices is the principle object of study. The empirical spectral distribution $F_{n^{-1}\X \X'}$ is constructed from the eigenvalues via
\begin{equation*}
F_{n^{-1}\X \X'}(x)= \frac{1}{p}\; \# \{ 1\le j\le p : n^{-1} \lambda_j \le x \}, \quad x\in \R,\quad n\ge1.
\end{equation*}
In the literature convergence results for the sequence of empirical spectral distributions are established under the assumption that $p$ and $n$ grow at the same rate.
Suppose that the iid entries $Z_{it}$ have mean $0$ and variance $1$. If \eqref{eq:gammach3} holds, then, with probability one, $(F_{n^{-1}\X \X'})$ converges weakly to the celebrated Mar\v cenko--Pastur law $F_\gamma$. If $\gamma \in (0,1]$,  $F_\gamma$  has density,
\begin{eqnarray}\label{eq:MPch3}
f_\gamma(x) =
\left\{\begin{array}{ll}
\frac{1}{2\pi x\gamma} \sqrt{(b-x)(x-a)} \,, & \mbox{if } a\le x \le b, \\
 0 \,, & \mbox{otherwise,}
\end{array}\right.
\end{eqnarray}\noindent
where $a=(1-\sqrt{\gamma})^2$ and $b=(1+\sqrt{\gamma})^2$. If $\gamma>1$, the \MP law is a mixture of a point mass at $0$ and the density function $f_{1/\gamma}$ with weights $1-1/\gamma$ and $1/\gamma$, respectively.  The point mass at $0$ is intuitively explained by the fact that, with probability $1$, $\min(p,n)$ eigenvalues $\lambda_i$ are non-zero. When $n=(1/\gamma) \; p$ and $\gamma >1$ one sees that the proportion of non-zero eigenvalues of the sample covariance matrix is $1/\gamma$ while the proportion of zero eigenvalues is $1-1/\gamma$.

While the finite second moment is the central assumption to obtain the Mar\v cenko--Pastur law as the limiting spectral distribution, the finite fourth moment plays a crucial role when studying the largest eigenvalues
\beam\label{eq:orderch3}
\la_{(1)}\ge \cdots \ge \la_{(p)}
\eeam
of $\X\X'$, where we suppress the dependence on $n$ in the notation.

\begin{figure}[htb!]
  \centering
  \subfigure[Standard normal entries]{
    \includegraphics[scale=0.34]{normal-TW.pdf}
  }
  \subfigure[Entry distribution: $\P(X=\sqrt{3}) = \P(X=-\sqrt{3} ) =
  1/6$, $\P(X=0)=2/3$. Note $\E X = 0$, { $\E[ X^2] = 1$, $\E [X^3] = 0$ and
  $\E[ X^4] = 3$, i.e., the first 4 moments of $X$ match those of the standard normal \ds .}]{
    \includegraphics[scale=0.34]{3point-TW.pdf}
  }
  \caption{Sample density function of the largest eigenvalue compared
    with the Tracy--Widom density function. The data matrix $ \X$ has
    dimension $200 \times 1000$. An ensemble of 2000 matrices is
    simulated.}
  \label{fig:normal-3point-TW}
\end{figure}

Assuming \eqref{eq:gammach3} and iid entries $X_{it}$ with zero mean, unit variance and finite fourth moment,  Geman~\cite{geman}
showed that
\beam\label{eq:gemanch3}
\dfrac {\la_{(1)}}{n} \stas \big(1+\sqrt{\gamma}\big)^2\,,\qquad \nto\,.
\eeam
Johnstone \cite{johnstone:2001} complemented this \slln\ by the corresponding \clt\ in the special case of iid standard normal
entries:
\beam\label{eq:tcch3}
 n^{2/3}\,\dfrac{(\sqrt{\gamma})^{1/3}}{\big(1+\sqrt{\gamma}\big)^{4/3}}\Big(\dfrac {\la_{(1)}}{n} -
\big(1+\sqrt{\tfrac pn }\big)^2\Big)
\std {\rm TW}\,,
\eeam
where the limiting \rv\ has a {\em Tracy--Widom \ds} of order 1. Notice that the centering
$\big(1+\sqrt{\tfrac pn }\big)^2$ can in general not be replaced by $(1+\sqrt{\gamma})^2$.
This \ds\ is ubiquitous in random matrix theory.
%It is defined via some ordinary differential equation; we refer to \cite{tracy:widom:2012} for a definition and properties.
Its distribution function $F_1$ is given by
\begin{equation*}
F_1(s) = \exp\Big\{
  -\frac{1}{2} \int_{s}^\infty [
    q(x) + (x - s) q^2(x)
 ] \dint x
\Big\}\,,
\end{equation*}
where $q(x)$ is the unique solution to the Painlev\'e II differential
equation
\begin{equation*}
  q''(x) = xq(x) + 2 q^3(x)\,,
\end{equation*}
where $ q(x)\sim {\rm Ai}(x)$ as $x \to \infty$ and Ai$(\cdot)$ is the Airy kernel; see Tracy and Widom~\cite{tracy:widom:2012} for details.
We notice that the rate $n^{2/3}$ compares favorably to the $\sqrt{n}$-rate in the classical \clt\ for sums
of iid finite variance \rv s.
The calculation of the spectrum is facilitated by the fact that the distribution of
the classical Gaussian matrix ensembles is invariant under orthogonal transformations. The corresponding
computation for non-invariant matrices with non-Gaussian entries is more complicated and was a major challenge for several years; a first step was made by Johansson \cite{johansson}.
Johnstone's result was extended to matrices $\X$ with iid non-Gaussian entries
by Tao and Vu \cite[Theorem~1.16]{tao09b}. Assuming that the first four moments of the entry \ds\ match  those of
the standard normal \ds , they showed \eqref{eq:tcch3} by
employing {\em Lindeberg's replacement method}, i.e., the iid non-Gaussian entries are replaced
step-by-step by iid Gaussian ones.
This approach is well-known from summation theory for \seq s of iid \rv s. Tao and Vu's result is a consequence of the so-called {\em Four Moment Theorem}, which describes the insensitivity of the eigenvalues with respect to changes in the distribution of the entries. To some extent (modulo the strong moment matching conditions) it shows the universality of Johnstone's limit
result \eqref{eq:tcch3}. Later we will deal with entries with infinite fourth moment. In this case, the weak limit
for the normalized largest eigenvalue $\la_{(1)}$ is distinct from the Tracy--Widom \ds : the classical Fr\'echet extreme value
\ds\ appears.
In Figure~\ref{fig:normal-3point-TW} we illustrate how the Tracy--Widom approximation works for Gaussian and non-Gaussian entries of $\X$
and in Figure~\ref{fig:MyDist-Frechet} we also illustrate that this approach fails when $\E[X^4]=\infty$.

Figure \ref{fig:normal-3point-TW} compares the sample density function
of the properly normalized largest eigenvalue estimated from 2000 simulated sample covariance matrices $\X\X'$ ($n=1000, p=200$) with
the Tracy--Widom density. If  $X$ has infinite fourth moment and further regularity conditions on the tail hold then the
Tracy--Widom limiting law needs to be replaced by the \Frechet
distribution; see Section~\ref{subsec:1.2} for details. Figure \ref{fig:MyDist-Frechet} illustrates this fact with a
simulated ensemble whose entries are distributed according to
the heavy-tailed distribution from \eqref{eq:distrsimch3} below with $\alpha = 1.6$.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5]{MyDist-Frechet.pdf}
  \caption{Sample distribution function of the largest eigenvalue $\la_{(1)}$
    compared to the \Frechet distribution (solid line) with $\alpha=1.6$. The data matrices have dimension
    $200 \times 1000$ and iid entries with infinite fourth moment. The results are based on 2000 replicates.}
  \label{fig:MyDist-Frechet}
\end{figure}

\subsection{The heavy-tailed case}\label{subsec:1.2}
So far we focused on ``light-tailed'' $\X$
in the sense that its entries have finite fourth moment.  However, there is statistical evidence that the assumption
of finite fourth moment may be violated when dealing with data from
insurance, finance or telecommunications. We illustrate this fact
in Figure~\ref{fig:SP500_tail_indices} where we show the pairs $(\alpha_L,\alpha_U)$ of
lower and upper tail indices
of $p=478$  log-return series composing
the S\&P 500 index estimated from $n=1,345$ daily observations from 01/04/2010 to 02/28/2015.
This means we assume for every row  $(X_{it})_{t=1,\ldots,n}$ of $\X$ that the tails behave like
\beao
\P(X_{it}>x)\sim c_U\,x^{-\alpha_U}\qquad\mbox{and}\qquad \P(X_{it}<-x)\sim c_L\,x^{-\alpha_L}\,,\qquad \xto\,,
\eeao
for non-negative constants $c_L,c_U$. We apply the Hill estimator (see Embrechts et al.~\cite{embrechts:kluppelberg:mikosch:1997}, p.~330,
de Haan and Ferreira \cite{dehaan:ferreira:2006}, p.~69)
to the \ts\ of the gains and losses in a naive way,
neglecting the dependence and non-stationarity in the data; we also omit confidence bands.
From the figure it is evident that the majority of the return series have tail indices below
four, corresponding to an infinite fourth moment.
\begin{figure}[htb!]
\centering
\includegraphics[scale=0.5]{SP500_tail_indices.pdf}
\caption{Tail indices of log-returns of 478 \ts\ from the S\&P 500 index. The values  $(\hat \alpha_L,\hat \alpha_U)$ of the
lower and upper tail indices are provided by
Hill's estimator.  We also draw the line $\hat\alpha_U=\hat\alpha_L$.}
\label{fig:SP500_tail_indices}
\end{figure}
The behavior of the largest eigenvalue $\la_{(1)}$
changes dramatically when $\X$ has infinite fourth moment.
Bai and Silverstein~\cite{baisilv} proved for an $n\times n$ matrix $\X$ with iid centered entries
that
\beam\label{eq:wdfrch3}
\limsup_{\nto} \dfrac{\la_{(1)}}{n}=\infty \qquad {\rm a.s.}
\eeam
This is in stark contrast to Geman's result \eqref{eq:gemanch3}.
%It is also unlikely that the
%Tracy--Widom \ds\ yields a good approximation to the \ds\ of the largest eigenvalue of the sample covariance matrix;
%see Figure~\ref{fig:MyDist-Frechet} for some simulation evidence.
\par
In the heavy-tailed case it is common to assume a {\em \regvar\ condition:}
\beam\label{eq:regvarch3}
\P(X>x)\sim p_+\,\dfrac{L(x)}{x^{\alpha}}\qquad\mbox{and}\qquad \P(X<-x) \sim p_-\,\dfrac{L(x)}{x^\alpha}\,,\qquad \xto\,,
\eeam
where $ p_\pm$ are non-negative constants \sth\ $p_++p_-=1$ and $L$ is a \slvary\ \fct . In particular, if $\alpha<4$ we have $\E [X^4]= \infty$. The \regvar\ condition
on $X$ (we will also refer to $X$ as a \regvary\ \rv ) is needed for proving \asy\ theory for
the eigenvalues of $\X\X'$. This is similar to proving limit theory for sums
of iid \rv s with infinite variance stable limits; see for example Feller~\cite{feller}.
\par

In \eqref{eq:MPch3} we have seen that the sequence $(F_{n^{-1}\X \X'})$ of empirical spectral distributions converges to the Mar\v cenko--Pastur law if the centered iid entries possess a finite second moment. Now we will discuss the situation when the entries are still iid and centered, but have an infinite variance. Here we assume the entries to be regularly varying with index $\alpha \in (0,2)$.
Assuming \eqref{eq:gammach3} with $\gamma\in (0,1]$ in this infinite variance case, Belinschi et al.~\cite[Theorem~1.10]{belinschi:dembo:guionnet:2009} showed that the sequence $(F_{a_{n+p}^{-2}\X \X'})$ converges with probability one to a non-random probability measure with density $\rho_{\alpha}^\gamma$ satisfying
\begin{equation*}
\rho_{\alpha}^\gamma(x) x^{1+\alpha/2} \to \frac{\alpha \gamma}{2(1+\gamma)}, \quad x \to \infty,
\end{equation*}
see also Ben Arous and Guionnet \cite[Theorem~1.6]{arous:guionnet:2008}.
The normalization $(a_k)$ is chosen \sth\ $\P(|X|>a_k)\sim k^{-1}$ as $k\to\infty$. An application of the Potter bounds (see Bingham et al. \cite[p.~25]{bingham:goldie:teugels:1987}) shows that $a_{n+p}^2/n \to \infty$.

It is interesting to note that there is a phase change in the extreme eigenvalues in going from finite to infinite fourth moment, while the phase change occurs for the empirical spectral distribution going from finite to infinite variance.
%In the case of finite fourth moment, \eqref{eq:geman} and \eqref{eq:MP} imply that $\lambda_{(k)}/n$ has to converge to the right endpoint of the Mar\v cenko--Pastur law for any finite $k$. However, if we assume an infinite fourth moment but finite second moment the empirical spectral distributions  $(F_{{n}^{-1}\X \X'})$ still converge to the Mar\v cenko--Pastur law with the same support, while $\lambda_{(1)}/n$ will tend to infintity due to \eqref{eq:wdfr}. This means that the proportion of the normalized eigenvalues taking values larger than $(1+\sqrt{\gamma})^2$ has to be negligible; otherwise we would have a contradiction to the convergence of the empirical spectral distributions to the Mar\v cenko--Pastur law.

%Finally, if the second moment is infinity, i.e.~regularly varying entries with index $\alpha \in (0,2)$, then the much larger values of the eigenvalues are visible in the limiting spectral distribution even when using a much stronger normalization \cite[Theorem~1.10]{belinschi:dembo:guionnet:2009}.
\par

The theory for the largest eigenvalues of sample covariance matrices with heavy tails is less developed than in the light-tailed case.
Pioneering work for $\la_{(1)}$ in the case of iid \regvary\ entries $X_{it}$ with index $\alpha\in (0,2)$
is due to Soshnikov~\cite{soshnikov:2004,soshnikov:2006}. He showed the \pp\ \con\
\beam\label{eq:nn}
N_n=\sum_{i=1}^p \vep_{a_{np}^{-2}\la_i} \std N=\sum_{i=1}^\infty \vep_{\Gamma_i^{-2/\alpha}}\,,\qquad \nto\,,
\eeam
under the growth condition \eqref{eq:gammach3} on $(p_n)$.
%The normalization $(a_k)$ is chosen \st\ $\P(|X|>a_k)\sim k^{-1}$ as $k\to\infty$.
Here
\beam\label{eq:Gamma}
\Gamma_i=E_1+\cdots + E_i\,,\qquad i\ge 1\,,
\eeam
and $(E_i)$ is an iid standard exponential \seq . In other words, $N$ is a Poisson point process on $(0,\infty)$ with mean \ms\
$\mu(x,\infty)= x^{-\alpha/2}$, $x>0$. Convergence in \ds\ of \pp es is understood in the sense of weak \con\
in the space of point \ms s equipped with the vague topology; see Resnick \cite{resnick:2007,resnick:1987}.
We can easily derive the limiting distribution of $a_{np}^{-2} \lambda_{(k)}$ for fixed $k\ge 1$ from \eqref{eq:nn}:
\begin{equation*}
\begin{split}
\lim_{\nto}\P(a_{np}^{-2} \lambda_{(k)}\le x)&= \lim_{\nto}\P(N_n(x,\infty)<k)
=  \P(N(x,\infty)<k) =\P(\Gamma_k^{-2/\alpha}\le x)\\ &= \sum_{s=0}^{k-1} \frac{\big(\mu(x,\infty)\big)^s}{s!} \e^{-\mu(x,\infty)}, \quad  x>0.
\end{split}
\end{equation*}
In particular,
\beao
\dfrac{\la_{(1)}}{a_{np}^2}\std \Gamma_1^{-\alpha/2}\,,\qquad \nto\,,
\eeao
where the limit has {\em Fr\'echet \ds } with parameter $\alpha/2$ and distribution function
\beao
\Phi_{\alpha/2}(x) =\ex^{-x^{-\alpha/2}}\,,\qquad x>0\,.
\eeao

\noindent We mention that the tail balance condition \eqref{eq:regvarch3} may be replaced in this case by the weaker assumption
$\P(|X|>x)= L(x) x^{-\alpha}$ for a \slvary\ \fct\ $L$. Indeed, it follows from the proof %(see also the comments below)
that only the squares $X_{it}^2$ contribute to the \pp\ limits of the eigenvalues $(\la_i)$. A con\seq\ of the
\cmt\ and \eqref{eq:nn} is the joint \con\ of the upper order statistics: for any $k\ge 1$,
\beao
a_{np}^{-2} \big(\la_{(1)},\ldots,\la_{(k)}\big)\std \big(\Gamma_1^{-2/\alpha},\ldots,\Gamma_k^{-2/\alpha}\big)\,,\qquad \nto\,.
\eeao

It follows from standard theory
for \pp es with iid points (e.g. Resnick \cite{resnick:2007,resnick:1987})
that \eqref{eq:nn} remains valid if
we replace  $N_n$  by the \pp\ $\sum_{i=1}^p\sum_{t=1}^n\vep_{X_{it}^2/a_{np}^2}$. Then we also have for any $k\ge 1$,
\beam\label{eq:nna}
a_{np}^{-2} \big(X_{(1),np}^2,\ldots,X_{(k),np}^2\big)\std \big(\Gamma_1^{-2/\alpha},\ldots,\Gamma_k^{-2/\alpha}\big)\,,\qquad \nto\,,
\eeam
where
\beao
X_{(1),np}^2\ge \cdots \ge X_{(np),np}^2
\eeao
denote the order statistics of $(X_{it}^2)_{i=1,\ldots,p;t=1,\ldots,n}$.
\par
Auffinger et al.~\cite{auffinger:arous:peche:2009} showed that \eqref{eq:nn}
remains valid under the \regvar\ condition \eqref{eq:regvarch3} for  $\alpha\in (2,4)$, the growth condition
\eqref{eq:gammach3} on $(p_n)$ and the
additional assumption $\E [X]=0$. Of course, \eqref{eq:nna} remains valid as well.
Davis et al.~\cite{davis:pfaffel:stelzer:2014} extended these results to the case when the rows of $\X$
are iid linear processes with iid \regvary\ noise. The Poisson \pp\ \con\ result of \eqref{eq:nn} remains valid in this case.
Different limit processes can only be expected if there is dependence across rows and columns.
\par
In what follows, we refer to the  {\em heavy-tailed case} when we assume the \regvar\ condition \eqref{eq:regvarch3}
for some $\alpha\in (0,4)$.%{\blue This hypothesis is confirmed in}

\subsection{Overview}

The primary objective of this overview  is to make a connection between extreme value theory and the behavior of the largest eigenvalues of sample covariance matrices from heavy-tailed multivariate time series.  For time series that are linearly dependent through time and across rows, it turns out that the extreme eigenvalues are essentially determined by the extreme order statistics from an array of iid random variables.  The asymptotic behavior of the extreme eigenvalues is then derived routinely from classical extreme value theory.  As such, explicit joint distributions of the extreme order statistics can be given which yield a plethora of ancillary results.  
%These include the limit behavior of ratios of consecutive extreme eigenvalues, ratios of extreme order statistics to the sample variance, etc..
%%The underlying multivariate \ts\ will assumed to be heavy-tailed and linearly dependent through time and across the rows.We consider asymptotic theory for the largest eigenvalues of a sample covariance matrix,
%and we will also address the largest singular values of the sample autocovariance matrix.
%To our knowledge, the present paper is the first to consider bonafide dependence among the components in the time series which renders a multivariate analysis, such as PCA, meaningful.
Convergence of the point process of extreme eigenvalues, properly normalized, plays a central role in establishing the main results.
\par
In Section~\ref{sec:2ch3} we continue the study of the case when the data matrix $\X$ consists of iid heavy-tailed entries.  We will consider power-law growth rates on the dimension $(p_n)$ that is more general than prescribed by \eqref{eq:gammach3}.
In Section~\ref{sec:model} we introduce a model for $X_{it}$ which allows for linear dependence across the rows and through time.
The point process convergence of normalized eigenvalues is presented in Section~\ref{sec:mainresult}.
This result lays the foundation for new insight into the spectral behavior of the sample covariance matrix, which is the content of Section~\ref{sec:samplecov}.

Sections~\ref{sec:samplecov} and \ref{sec:possemidef} are devoted to {\em sample autocovariance matrices}.  Motivated by \cite{lam:yao}, we study the eigenvalues of sums of transformed matrices and illustrate the results in two examples. These results are applied to the time series of S\&P 500 in Section \ref{sec:sp500}.
%The technical proofs are collected in Section~\ref{sec:proof}. 
Appendix~\ref{appendix:Ach3} contains useful facts about regular variation and point processes.

\section[General growth rates for $p_n$]{General growth rates for $p_n$ in the iid heavy-tailed case}\label{sec:2ch3}
This section is based on ideas in Heiny and Mikosch \cite{heiny:mikosch:2015:iid} where one can also find detailed proofs.
\subsection*{Growth conditions on $(p_n)$}\label{subsec:pn}
In many applications it is not realistic to assume
that the dimension $p$ of the data and the sample size $n$ grow at the same rate.
The aforementioned results of Soshnikov~\cite{soshnikov:2004,soshnikov:2006} and Auffinger et al. \cite{auffinger:arous:peche:2009} already indicate that the value $\gamma$ in the growth
rate~\eqref{eq:gammach3} does not appear in the \ds al limits.
This obervation is in contrast to the light-tailed case; see \eqref{eq:gemanch3}
and \eqref{eq:tcch3}.  Davis et al.~\cite{davis:mikosch:pfaffel:2015,davis:pfaffel:stelzer:2014}
and Heiny and Mikosch ~\cite{heiny:mikosch:2015:iid} allowed for more general rates for
$p_n\to\infty$ than linear growth in $n$.
Recall that $p=p_n\to\infty$ is the number of rows in the
matrix $\X_n$. We need to specify the growth rate of $(p_n)$ to ensure
a non-degenerate limit distribution of the normalized singular values of the sample autocovariance
matrices. To be precise, we assume
\begin{equation}\label{eq:pch3}
p=p_n=n^\beta \ell(n), \quad n\ge1,\tag{$C_p(\beta)$}
\end{equation}
where $\ell$ is a slowly varying function and $\beta\ge 0$. If $\beta =0$, we also assume $\ell(n) \to \infty$.
Condition~\ref{eq:pch3} is more general than the growth conditions in the literature; see
\cite{auffinger:arous:peche:2009,davis:mikosch:pfaffel:2015,davis:pfaffel:stelzer:2014}.
\begin{theorem}\label{thm:intro}
Assume that $\X=\X_n$ has iid entries satisfying the \regvar\ condition \eqref{eq:regvarch3} for some
$\alpha \in (0,4)$. If $\E[|X|]<\infty$ we also suppose that $\E [X]=0$. Let $(p_n)$ be an integer sequence
satisfying \ref{eq:pch3} with $\beta\ge 0$. In addition, we require
\begin{equation}\label{Cbeta}
\min(\beta,\beta^{-1})\in (\alpha/2-1,1] \qquad \mbox{ for } \alpha\in [2,4), \tag{$\widetilde{C}_\beta(\alpha)$}
\end{equation}
Then
\beam\label{eq:2}
\sum_{i=1}^p \vep_{a_{np}^{-2}\la_i}\std \sum_{i=1}^\infty \vep_{\Gamma_i^{-2/\alpha}}\,,\qquad \nto\,,
\eeam
where the \con\ holds in the space of point \ms s with state space $(0,\infty)$ equipped with the vague toplogy; see Resnick \cite{resnick:2007}.
\end{theorem}

\subsubsection*{A discussion of the case $\beta\in [0,1]$}
We mentioned earlier that in the heavy-tailed case, limit theory for the
largest eigenvalues of the sample covariance matrix is rather insensitive to
the growth rate of $(p_n)$ and that the limits are essentially
determined by the diagonal of this matrix. This is confirmed by the following result.

\begin{proposition}\label{prop:offdiagonalch3}
Assume that $\X=\X_n$ has iid entries satisfying the \regvar\ condition \eqref{eq:regvarch3} for some
$\alpha \in (0,4)$. If $\E[|X|]<\infty$ we also suppose that $\E [X]=0$. Then for any \seq\ $(p_n)$
satisfying \ref{eq:pch3} with $\beta\in [0,1]$ we have
\beao
a_{np}^{-2} \twonorm{\X \X' - \diag(\X \X')}\stp 0\,,\qquad\nto\,,
\eeao
where $\| \cdot\|_2$ denotes the spectral norm; see \eqref{specnorm} for its definition.
\end{proposition}

\par
Proposition~\ref{prop:offdiagonalch3} is not unexpected for two reasons:
\begin{itemize}
\item
It is well-known from classical theory (see Embrechts and Veraverbeke \cite{embrechts:veraverbeke:1982}) that for any iid \regvary\ non-negative \rv s
$Y,Y'$ with index  $\alpha'>0$, $Y\,Y'$ is \regvary\ with index $\alpha'$ while $Y^2$ is \regvary\ with index $\alpha'/2$.
Therefore $X^2$ and $X_{11}X_{12}$  are \regvary\ with indices $\alpha/2$ and $\alpha$, respectively.
\item
The aforementioned tail behavior is inherited by the entries of $\X\X'$ in the following sense.
By virtue of Nagaev-type \ld\ results for an iid \regvary\ \seq\ $(Y_i)$ with index $\alpha'\in (0,2)$ where we also assume that
$\E [Y_0]=0$ if $\E[|Y_0|]<\infty$
(see Theorem~\ref{thm:nagaevch3}) %S.V. Nagaev \cite{nagaev:1979} and
%Cline and Hsing \cite{cline:hsing:1998})
we have that $\P(Y_1+\cdots +Y_n >b_n)/(n \,\P(|Y_0|>b_n))$ converges to a non-negative constant
provided $b_n/a_n'\to\infty$, where $\P(|Y_0|>a_n')\sim n^{-1}$ as $\nto$. As a con\seq\ of the tail behaviors of $X_{it}^2$ and $X_{it}X_{jt}$ for $i\ne j$
and Nagaev's results we have for $(b_n)$ \sth\ $b_n/a_n^2\to\infty$,
\beam \label{eq:dfsdfjl}
\dfrac{\P\big((\X\X')_{ij}> b_n\big)}{\P\big((\X\X')_{ii}- c_n> b_n\big)}\sim \dfrac{n\,\P( X_{11}X_{12}>b_n)}{n\,\P(X^2>b_n)}\to 0\,,\qquad \nto\,,
\eeam
where $c_n=0$ or $n\,\E[X^2]$ according as $\alpha\in (0,2)$ or $\alpha\in (2,4)$.
This means that the diagonal and off-diagonal entries of
$\X\X'$ inherit the tails of $X_{it}^2$ and $X_{it}X_{jt}$, $i\ne j$, respectively, above the high threshold $b_n$.
\end{itemize}
\par
Proposition~\ref{prop:offdiagonalch3} has some immediate con\seq s for the approximation of the eigenvalues
of $\X\X'$ by those of ${\rm diag}(\X\X')$. Indeed, let $C$ be a symmetric $p\times p$ matrix with
eigenvalues $\la_1(C),\ldots,\la_p(C)$ and
ordered eigenvalues
\beam \label{eq:weyl}
\la_{(1)}(C)\ge \cdots \ge \la_{(p)}(C)\,.
\eeam
Then for any symmetric $p\times p$ matrices $A,B$, by {\em Weyl's inequality} (see Bhatia \cite{bhatia:1997}),
\beao
\max_{i=1,\ldots,p}\big|\la_{(i)}(A+B)-\la_{(i)}(A)\big|\le \|B\|_2\,.
\eeao
If we now choose
$A+B=\X\X'$ and $A= \diag (\X\X')$ we obtain the following result.

>>>>

\input{../papers/DavisMikoschHeinyXie/PaperBody}
